---
title: "Data Sonification of Endangered Languages"
author: "Alex Lu"
date: "5/8/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Introduction 

This semester, I had the honor of taking a linguistics class related to endangered languages,
a stats class where we programmed in R, and a music class that connects music to data. Thus, 
with this final project I thought it would be cool to use something related to all three classes.
From my linguistics class we learned that of the 7,000 endangered languages in existence today,
around 3,000 of them are endangered. The data related to these 3,000 languages are used as the 
dataset for this project. From my stats class, I'll be using R in order to process my data
and create my data visualizations. From this class, I plan on creating a data sonification
that turns data into sounds through TwoTones.

### 1) Importing necessary packages to create data visualization. 

For my project, I decided to use R in order to create my data visualizations. In order to do this
I relied on the following packages in the code block below. The "tidyverse" and "maps" packages are necessary to create
the scatterplots while "gganimate" is used to add animations to the generated scatterplot.

```{r,message=FALSE}
library(tidyverse)
library(gganimate)
require(maps)
library("av")
```

### 2) Import Endangered Languages data 

The data used for this project was published by "The Guardian" and contains a list 
of endangered languages and different columns including the different names, ID number,
country codes, latitude, longitude, etc. For the purposes of the project we only need 
the latitude,longitude, degree of endangerment, and name in English (although this isn't 
that important either). In the code below, all the extraneous columns are removed and the 
data is arranged in order of longitude.

```{r}
data<- read.csv(file='data.csv')
sorted_data<-data %>% 
  arrange(Longitude) %>% 
  select(Name.in.English,Degree.of.endangerment,Latitude,Longitude)
```

If we were to plot all the data points now and color code each point by the degree of endangerment
we'd get the following plot.

```{r, warning=FALSE}
theme_set(
  theme(panel.background= element_rect(fill="white"),legend.position="none")
)
world_map <- map_data("world")
ggplot(world_map, aes(x = long, y = lat,group=group)) +
  geom_polygon(fill="lightgray", colour = "darkgray")+
  geom_point(data=sorted_data,aes(x=Longitude,y=Latitude,group=Degree.of.endangerment
                                  ,color=Degree.of.endangerment))
```

For this project, I decided it'd be better if the data were separated depending on the 
degree of endangerment. This makes the plots less clustered. Furthermore, given that 
TwoTones has a max bpm of 300, the final audio file would be super long which means the 
animation would take a couple of minutes which would take an extremely long time to render.

### 3) Creating CSVs for each Degree of Endangerment 

In order to turn the data into sound, I need to create CSV files which will then be 
used in TwoTones to produce an audiofile with the sonified version of the data. For the 
purposes of this document, the "write.csv" commands have been commented out.

```{r} 

#Data for all "Critically Endangered" languages
critical_data<- sorted_data %>% 
  filter(Degree.of.endangerment=="Critically endangered")
critical_data<-na.omit(critical_data) 
#write.csv(critical_data,"critical_data.csv") 

#Data for all "Extinct" languages
extinct_data<- sorted_data %>% 
  filter(Degree.of.endangerment=="Extinct")
extinct_data<-na.omit(extinct_data) 
#write.csv(extinct_data,"extinct_data.csv") 

#Data for all "Vulnerable" languages 
vulnerable_data<- sorted_data %>% 
  filter(Degree.of.endangerment=="Vulnerable")
vulnerable_data<-na.omit(vulnerable_data) 
#write.csv(vulnerable_data,"vulnerable_data.csv") 

#Data for all "Severely endangered" languages 
severe_data<- sorted_data %>% 
  filter(Degree.of.endangerment=="Severely endangered")
severe_data<-na.omit(severe_data) 
#write.csv(severe_data,"severe_data.csv")
``` 

Using TwoTones is just a simple matter of dragging and dropping each csv file into the program
and choosing which column to use as the datasource (Latitude). From then I tweaked the timing of the 
data to play at 300bpm. This an extremely fast tempo but it's necessary for me in order to ensure
that the animations have a shorter duration which drastically decreases the amount of time it would take
for my computer to render each animation. 

### 4) Creating animations

Now that all the  data has been separated and processed, we can use "gganimate" to create
animations for each scatterplot. Because animations can't be played in a pdf file, the following
is what the scatter plots look like  

1) Extinct Plot 

```{r,echo=FALSE}
ggplot(world_map, aes(x = long, y = lat,group=group)) +
  geom_polygon(fill="lightgray", colour = "darkgray")+
  geom_point(data=extinct_data,aes(x=Longitude,y=Latitude,group=Degree.of.endangerment),color="red")
```

2) Severely Endangered Plot 

```{r, echo=FALSE}
ggplot(world_map, aes(x = long, y = lat,group=group)) +
  geom_polygon(fill="lightgray", colour = "darkgray")+
  geom_point(data=severe_data,aes(x=Longitude,y=Latitude,group=Degree.of.endangerment),color="orange")
``` 

3) Vulnerable Plot 

```{r, echo=FALSE} 
ggplot(world_map, aes(x = long, y = lat,group=group)) +
  geom_polygon(fill="lightgray", colour = "darkgray")+
  geom_point(data=vulnerable_data,aes(x=Longitude,y=Latitude,group=Degree.of.endangerment),color="green")

``` 

4) Critically Endangered Plot 

```{r, echo=FALSE}
ggplot(world_map, aes(x = long, y = lat,group=group)) +
  geom_polygon(fill="lightgray", colour = "darkgray")+
  geom_point(data=critical_data,aes(x=Longitude,y=Latitude,group=Degree.of.endangerment),color="blue")

```

#### Animation Code 

This code is used to create the actual animations for the plots shown above. This was by far 
the most time consuming part of the project due to the amount of time it takes for each
animation to render. The animations gradually add each data point into the scatter plot. 
The goal here is to sync each point with their respective note in the audio file. 

Note: Yes, the code is repetitive and I definitely could have used a for loop to shorten
down the code a lot. 

```{r,eval=FALSE} 
#Animation for "Critically endangered" languages
critical_plot<-ggplot(world_map, aes(x = long, y = lat,group=group)) +
  geom_polygon(fill="lightgray", colour = "darkgray")+
  geom_point(data=critical_data,aes(x=Longitude,y=Latitude,group=Degree.of.endangerment)
             ,color="blue") + transition_states(Longitude,state_length=0.1
             ,transition_length=0) + shadow_mark(color="blue",past=T,future=F,alpha=0.7)

d<-animate(critical_plot,height=1050,width=1680,res=200,renderer=av_renderer()
             ,duration=121,fps=20) 

#Animation for "Extinct" languages
extinct_plot<-ggplot(world_map, aes(x = long, y = lat,group=group)) +
  geom_polygon(fill="lightgray", colour = "darkgray")+
  geom_point(data=extinct_data,aes(x=Longitude,y=Latitude,group=Degree.of.endangerment)
             ,color="red") + transition_states(Longitude,state_length=0.1
             ,transition_length=0) + shadow_mark(color="red",past=T,future=F,alpha=0.7) 

b<-animate(extinct_plot,height=1050,width=1680,res=200,renderer=av_renderer()
             ,duration=50,fps=20) 

#Animation for "Severely Endangered" languages
severe_plot<-ggplot(world_map, aes(x = long, y = lat,group=group)) +
  geom_polygon(fill="lightgray", colour = "darkgray")+
  geom_point(data=severe_data,aes(x=Longitude,y=Latitude,group=Degree.of.endangerment)
             ,color="orange") + transition_states(Longitude,state_length=0.1
             ,transition_length=0) + shadow_mark(color="orange",past=T,future=F,alpha=0.7)

c<-animate(severe_plot,height=1050,width=1680,res=200,renderer=av_renderer()
             ,duration=110,fps=20)

#Animation for "Vulnerable" languages
vulnerable_plot<-ggplot(world_map, aes(x = long, y = lat,group=group)) +
  geom_polygon(fill="lightgray", colour = "darkgray")+
  geom_point(data=vulnerable_data,aes(x=Longitude,y=Latitude,group=Degree.of.endangerment)
             ,color="green") + transition_states(Longitude,state_length=0.1
             ,transition_length=0) + shadow_mark(color="green",past=T,future=F,alpha=0.7)

e<-animate(vulnerable_plot,height=1050,width=1680,res=200,renderer=av_renderer()
             ,duration=125,fps=20) 

#Saving all animations to files

anim_save("extinctplot.mp4",b)
anim_save("severeplot.mp4",c)
anim_save("criticalTest.mp4",d)
anim_save("vulnerableplot.mp4",e)
```  

### 5) Combining Audio and Animations 

Now that we have our animations and our audio files, we can combine them using some video editing 
software like iMovie to create our final data sonification product. There are going to be some 
inconsistencies between the visual and audio components which is unfortunate. Part of these issues are 
most likely due to the timing of the animation being slightly off. Out of the four sonifications
that were created, I think the sonification related to "Extinct" languages came out the best 
and was the most in sync.






